---
title : "R Subsetting"
date : "19 November 2015"
author : "Murray Cadzow"
output : html_document
---



In this lesson we will cover subsetting your data using:

- logical operations

- subset()

- square brackets

- intro dplyr

## Logical operations

Before we can subset we need to know how to create the conditions for R to know what to select

We do this through logical comparisions

The logical operators as mentioned before are:

- == ( equality  )

- != ( not equal )

- \> (greater than)

- \>= (greater than or equal to)

- < (less than)

- <= (less than or equal to)

- ! (not)

- \| (or)

- & (and)

The result of using these logical comparisons is a vector of **TRUE**s and **FALSE**s

Here are some examples:
```{r, eval = TRUE, echo = TRUE}
data("iris")
iris$Sepal.Length
iris$Sepal.Length < 5

# using not
!TRUE
!FALSE
!( iris$Sepal.Length < 5)

# equality
TRUE == TRUE
TRUE == FALSE
FALSE == FALSE
iris$Sepal.Length == 5

# not equality
TRUE != TRUE
TRUE != FALSE
FALSE != FALSE
iris$Sepal.Length != 5

# combine conditions or ( | )
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
iris$Sepal.Length > 5 | iris$Sepal.Length < 5

#combine conditions and ( & )
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
iris$Sepal.Length > 5 & iris$Petal.Width < 1.3

# different types
3 < 'a'
```

**NOTE** comparisons will be performed on the "lowest common data type" which is usually _character_ when different types are involved

## _Subset()_

The subset function provides a easy way to subset your data into groups based on your logical conditions

We'll use the dummy data for these examples

```{r, echo = TRUE, eval = TRUE}
data("iris")

# what does it look like?
dim(iris)
head(iris)

# select Sepal lengths under 5
under5 <- subset(x = iris, iris$Sepal.Length < 5)

# how many people do we now have?
dim(under5)

# from our original data how many people are male (SEX == 1) and have gout (GOUTAFFSTAT == 2)
setosaSepalLengthOver5 <- subset(x = iris, Species == 'setosa' & Sepal.Length > 5)
dim(setosaSepalLengthOver5)

```

### Subset with square brackets

Just as we can use square brackets and indices to select items we want, instead of indices we can provide a logical condition


```{r, echo = TRUE, eval = TRUE}
# repeat under 5 selection
# here we're saying we want all the columns but select the rows where the condition is met
under5_repeat <- iris[iris$Sepal.Length < 5 , ]
dim(under5_repeat)

```


### Intro to dplyr

Made to make data manipulation easy

Uses verb functions to make thoughts -> code easier

We'll focus on the subsetting functions to start with

```{r, eval=TRUE, echo = TRUE}
library('dplyr')

# filter rows with filter()
head( filter(iris, Sepal.Length < 5) )

# or slice()
slice(iris, 1:10)

# select columns with select()
head( select(iris, Sepal.Length, Species) )

```

Pipe %>%

Instead of wrapping functions with dplyr you can pipe the output of a function into the input of another

```{r, eval = TRUE, echo = TRUE}
iris %>% select(Sepal.Length) %>% head()
```

We can also create summaries and group data
```{r, eval = TRUE, echo = TRUE}
iris %>% summarise(mean(Sepal.Length))

iris %>% group_by(Species) %>% summarise(mean(Sepal.Length))
```
